---
title: |
  An automated workflow to standardise taxon names for South African alien species lists

author:
  - Katelyn T. Faulkner$^{1,2}$

  - $^1$South African National Biodiversity Institute, Kirstenbosch Research Centre, Cape Town, 
  - 7735, South Africa
  - $^2$Department of Zoology and Entomology, University of Pretoria, Pretoria, 0028, 
  - South Africa
date: "5 February 2025"

output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE, cache = FALSE}
knitr::opts_knit$set(
  root.dir = rprojroot::find_rstudio_root_file()
  )
```


```{r global_options, include = FALSE}
knitr::opts_chunk$set(message=FALSE, warning = FALSE,
tidy.opts=list(width.cutoff=60, tidy=TRUE), eval=FALSE) 
```

\newpage

# Introduction

This automated workflow was developed to standardise the names of alien taxa for South African alien species lists. It can be used to standardise digitised lists of taxon names that are in one of two formats: (1) full scientific name with authorship and date information (hereafter ‘scientific name’); and (2) scientific name without authorship and date information (hereafter ‘canonical name’). While the workflow can be used to taxonomically standardise canonical names, it is recommended that information on authorship and date is retained as this decreases errors and makes it easier to resolve issues. The workflow implements taxonomic backbones that are relevant to the South African context.

Three taxonomic backbones are implemented: that of the Global Biodiversity Information Facility (GBIF); that of the Botanical Database of Southern Africa (BODATSA); and that of The World Checklist of Vascular Plants (WCVP), accessed via Plants of the World Online (POWO) (https://powo.science.kew.org/).

The taxonomic backbone used depends on the taxonomic group:

1. The GBIF backbone is used for non-plant taxa (e.g., animals, fungi, and chromista)

2. The BODATSA backbone is used for alien plants recorded outside of captivity or cultivation in South Africa (i.e., those taxa included in BODATSA)

3. The WCVP backbone is used for alien plants not found outside of captivity or cultivation in South Africa (i.e., those taxa not included in BODATSA)

In addition to standardising a list of taxon names based on the three taxonomic backbones, the workflow obtains canonical names and higher taxonomic information for the taxa from GBIF, and detects, corrects, and flags potential issues and errors. 

This document provides the script for the workflow with annotations.

Further details are available at the GitHub repository for the workflow: "https://github.com/KatelynFaulkner/rsa-ans-workflow"

The workflow has been created in R software, version 4.4.0 (R Core Team 2024). 

\newpage

# Step 1: Preparation of the R environment

#### Install and load R CRAN packages

```{r}
packages = c("tidyr", "dplyr",
             "rgbif", "stringr", "stringdist", "rWCVP", "purrr") # seven packages are required
```

```{r}
package.check <- lapply(
  packages,
  FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
      library(x, character.only = TRUE)
    }
  }
)
```

#### Install and load "remote" package from R CRAN, and install and load "rWCVPdata" package from GitHub

```{r}
remotes.check<-"remotes" %in% rownames(installed.packages()) 
if(remotes.check == FALSE){
  install.packages("remotes")
  remotes::install_github('matildabrown/rWCVPdata') # one package required from GitHub
}
require('rWCVPdata') # load package from GitHub
```

# Step 2: Preparation of data

## Step 2a: Preparation of BODATSA dataset

#### Load dataset from BODATSA

```{r}
BODATSA<-read.csv("data\\raw\\BODATSA.csv") # file must be named "BODATSA.csv"
```

#### Provide required information on BODATSA

```{r}
BODATSADate<-"02 December 2024" # stipulate the date BODATSA was downloaded in the format shown here
```

#### View header of data 

```{r}
head(BODATSA) 
```

#### Assign NA to records where data are missing

```{r}
BODATSA[BODATSA == ""] <- NA 
```

#### In "Rank1" column replace "ssp." with "subsp."

```{r}
BODATSA<-BODATSA %>%
  mutate(Rank1=replace(Rank1, Rank1=="ssp.", "subsp."))
```

#### Add a column with the canonical names of the taxa to BODATSA dataset

```{r}
BODATSA <- BODATSA %>% unite("canonicalName", c(Genus, Sp1, Rank1, Sp2, 
                                                Rank2, Sp3), na.rm = TRUE, sep = " ", remove = FALSE)
```

#### Add a column with the scientific names of the taxa to BODATSA dataset

```{r}
BODATSA <- BODATSA %>% unite("scientificName", c(Genus, Sp1, Author1, Rank1, Sp2, Author2, Rank2, Sp3, Author3), na.rm = TRUE, sep = " ", remove = FALSE)
```

#### Correct and flag incorrectly formatted names in the "Accepted" column

Identify the names in the "Accepted" column that are incorrectly formatted

```{r}
FormatIssue<-BODATSA%>% filter(!Accepted %in% BODATSA$scientificName & !is.na(Accepted))
```

Identify the accepted name in the correct format using fuzzy matching

```{r}
fullScientificName<-{} # create empty dataset
for(i in 1:length(FormatIssue$Accepted)){ # loop through names with format issues
  keywords<-unlist(str_split(FormatIssue$Accepted[i], " ")) # split name
  y<-paste(keywords[1:2], collapse = " ") # create string that only contains the genus and species of the name
  z <-  BODATSA %>%
    filter(str_detect(scientificName, y)) # subset BODATSA to only include records for that genus and species
  a<-stringdist(FormatIssue$Accepted[i], z$scientificName, method="jaccard") # fuzzy match the name to those of the genus and species with jaccard method
  b<-which(a==min(a)) # get the closest match
  if(length(b)==1){
  c<-z$scientificName[b] # take most similar match if there is only one
  }
  if(length(b)>1){ # if more than one closest match
   issues<-data.frame(z$scientificName[b]) 
  names(issues)<-"scientificName"
   issues<-issues %>%
      mutate(text = str_split(scientificName, '\\s+'), # split the names of the potential matches
             words_correct = map_dbl(text, ~sum(.x %in% keywords)))
    d<-which(issues$words_correct == max(issues$words_correct)) # determine number of components in the potential matches that are equal to those in the taxon name
    if(length(d) == 1){
    c<-issues$scientificName[d] # take the one with the highest number of equal components
  }
  else if(length(d)> 1){ # if more than one option with the highest number of equal components
    g<-stringdist(FormatIssue$Accepted[i], issues$scientificName[d], method="lv") # fuzzy match using lv method
    h<-which(g==min(g)) # identify the closest match
    c<-issues$scientificName[d][h] # take the one that is the closest match
  }
  }
  fullScientificName<-rbind(fullScientificName, c) # bind the name identified through fuzzy matching to result data frame
}
```

```{r}
FormatIssue<-cbind(FormatIssue,fullScientificName) # bind the matched names to those with format issues
```

Replace incorrectly formatted names in "Accepted" column with those identified through fuzzy matching, and flag

```{r}
FormatIssueAcc<-FormatIssue %>% select(Accepted, fullScientificName) # remove unnecessary columns
FormatIssueAcc<-unique(FormatIssueAcc) # remove duplicated rows

BODATSA <- left_join(BODATSA, FormatIssueAcc, by = c("Accepted" = "Accepted")) # join the matched names to BODATSA

BODATSA <- BODATSA %>% 
  rename("AcceptedIssues"= "Accepted") # rename the original "Accepted" column "AcceptedIssues"
```

Create a new "Accepted" column with the matched names 

```{r}
BODATSA<-BODATSA %>%
  mutate(Accepted = replace(AcceptedIssues, is.na(fullScientificName)==FALSE, fullScientificName[is.na(fullScientificName)==FALSE]))
```

Create a column indicating where changes were made

```{r}
BODATSA<-BODATSA %>%
  mutate(synAccIssue = ifelse(!is.na(fullScientificName), TRUE, FALSE)) # column is named "synAccIssues"
```

Delete unnecessary column

```{r}
BODATSA$fullScientificName<-NULL
```

#### Write to file

```{r}
write.csv(BODATSA, "data//interim//BODATSAPrepared.csv", row.names = FALSE)
```

## Step 2b: Preparation of list of taxon names

#### Load list of names for standardisation

```{r}
NamesDat<-read.csv("data\\raw\\OriginalNames.csv") # file must be named "OriginalNames.csv"
```

#### Provide required information on list of names for standardisation

```{r}
NameType<-"canonical" # here stipulate the format of the taxon names: 'canonical' or 'scientific'
```

#### View header of data and get the number of taxa

```{r}
head(NamesDat) # Look at the first five rows of the data.

length(NamesDat$verbatimScientificName) # how many taxa are in the dataset
```

#### Standardise the spacing formats 

```{r}
NamesDat<-data.frame(verbatimScientificName = str_replace_all(
  NamesDat$verbatimScientificName, "\\s", " ")) # standardise spacing formats

NamesDat$verbatimScientificName<-str_squish(NamesDat$verbatimScientificName) # remove double spaces
```

# Step 3: Match all taxon names to GBIF backbone

#### Match all names to GBIF backbone, and get canonical names and higher taxonomic information

```{r}
GBIFCnames<-c("usageKey","scientificName", "canonicalName", "rank",          
              "status", "confidence", "matchType", "kingdom",       
              "phylum", "order", "family", "genus",         
              "species", "kingdomKey", "phylumKey", "classKey",      
              "orderKey", "familyKey", "genusKey", "speciesKey",    
              "synonym", "class") # create a list of ideal column names for the GBIF data
```

```{r}
GBIFMatch<-{} # create empty dataframe for the results
m<-length(NamesDat$verbatimScientificName)
for(j in 1:m){ # loop through the taxon names and check each taxon name against the GBIF backbone
  x<-data.frame(name_backbone(NamesDat$verbatimScientificName[j])) # get information from GBIF for that name
  x[setdiff(GBIFCnames, names(x))] <- NA 
  x<-x[GBIFCnames]
  if(is.na(x$family)){ # if the name was not matched, bind the required data
    u<-cbind(NamesDat[j,], x$canonicalName, x$kingdom, x$phylum, x$class, x$order,x$family, x$scientificName, x$synonym, x$rank, x$status, x$matchType, x$usageKey)
  }
  else if(x$status == "ACCEPTED"|x$status == "DOUBTFUL"){ # if the name is accepted or doubtful, bind the required data
    u<-cbind(NamesDat[j,], x$canonicalName, x$kingdom, x$phylum, x$class, x$order,x$family, x$scientificName, x$synonym, x$rank, x$status, x$matchType, x$usageKey)
  }
  else if(x$status == "SYNONYM"){ # if name is a synonym, fetch the accepted scientific name, and bind required columns
    a<-name_backbone(x$species, kingdom = x$kingdom)
    u<-cbind(NamesDat[j,], x$canonicalName, x$kingdom, x$phylum, x$class, x$order,x$family, a$scientificName, x$synonym, x$rank, x$status, x$matchType, x$usageKey)
  }
  colnames(u)[c(ncol(u)-12, ncol(u)-11, ncol(u)-10, ncol(u)-9, ncol(u)-8, ncol(u)-7, ncol(u)-6, ncol(u)-5, ncol(u)-4, ncol(u)-3,ncol(u)-2, ncol(u)-1,
                ncol(u))]<-c("verbatimScientificName", "canonicalName", "kingdom", "phylum", "class", "order","family", "scientificName", "synonym", "rank", "status","matchtype", "usageKey") # name the columns
  GBIFMatch<-as.data.frame(rbind(GBIFMatch,u)) # bind data to result dataframe
}
```

#### Write interim output to file 

```{r}
GBIFInterim<-GBIFMatch

write.csv(GBIFInterim, "data//interim//GBIFInterim.csv", row.names = FALSE)
```

# Step 4: Extract plants

```{r}
unique(GBIFMatch$kingdom) # list of kingdoms in the output
```

#### Create a plants subset

```{r}
PlantDat<-GBIFMatch %>% filter(kingdom == "Plantae")  # create subset of plants

PlantDat<-PlantDat %>% select(verbatimScientificName:family) # only keep the relevant columns
names(PlantDat) # names of the columns
```

#### Create a non-plants subset

```{r}
GBIFMatch<-GBIFMatch %>% filter(kingdom != "Plantae"|is.na(kingdom)) # create a non-plants subset 
```

# Step 5 Assess and flag issues for non-plant taxa

#### Assess issues

```{r}
head(GBIFMatch) # look at the first 5 rows of the dataset
```

Look at how many have each status (e.g., accepted, synonym etc)

```{r}
GBIFMatch %>%
  group_by(status) %>%
  count()
```

Look at how many had a specific match type (e.g., exact, fuzzy etc)

```{r}
GBIFMatch %>%
  group_by(matchtype) %>%
  count()
```

Look at how many fall into different ranks (e.g., species, subspecies etc)

```{r}
GBIFMatch %>%
  group_by(rank) %>%
  count()
```

#### Flag taxonomic issues

Flag synonyms

```{r}
GBIFMatch <- GBIFMatch %>% 
  rename("synonym.flag"= "synonym") # give TRUE if a synonym
```

Flag taxon names that were not exact matches (i.e., fuzzy matches)

```{r}
GBIFMatch = GBIFMatch %>% 
  mutate(uncertain.flag=case_when((matchtype =="FUZZY")~TRUE,
                                  TRUE~FALSE)) # give TRUE if a fuzzy match
```

Flag doubtful taxon names (e.g., taxonomic homonyms, missapplied taxon names)

```{r}
GBIFMatch = GBIFMatch %>% 
  group_by(status) %>% 
  mutate(doubtful.flag = ifelse(status =="DOUBTFUL" & !is.na(status), TRUE, FALSE)) # give TRUE if doubtful
```

Flag unmatched taxa (those not matched to an accepted name, or matched at higher levels (i.e., genus or above level for a species))

```{r}
GBIFMatch = GBIFMatch %>% 
  mutate(unmatched.flag=case_when((matchtype =="HIGHERRANK")~TRUE,
                                     (rank == "GENUS")~TRUE,
                                     (matchtype == "NONE") ~TRUE,
                                     TRUE~FALSE)) # give TRUE if not matched
```

Flag if the result should be checked for a possible error or issue (unmatched, doubtful, and uncertain matches)

```{r}
GBIFMatch<-GBIFMatch %>% 
  mutate(possibleError.flag = ifelse(uncertain.flag =="TRUE" | doubtful.flag == "TRUE" | unmatched.flag == "TRUE", TRUE, FALSE)) # TRUE if there could be an issue
```

Flag sub-specific entities

```{r}
GBIFMatch = GBIFMatch %>% 
  group_by(rank) %>% 
  mutate(subspecific.flag = ifelse(rank =="SPECIES"|rank == "GENUS"|is.na(rank)|matchtype == "HIGHERRANK", FALSE, TRUE)) # TRUE if sub-specific
```

Flag duplicates in returned scientific names (exclude duplicates if name is at higher taxonomic levels (e.g., genus))

```{r}
GBIFMatch = GBIFMatch %>% 
  group_by(scientificName) %>% 
  mutate(scientificNameDuplicate.flag = ifelse(n() > 1 & rank !="GENUS" & rank !="KINGDOM" & rank !="PHYLUM" & rank !="CLASS" & rank != "ORDER" & rank != "FAMILY" & !is.na(rank),
                                               TRUE, FALSE)) # TRUE if a duplicate
```

Flag duplicates in verbatim names 

```{r}
GBIFMatch = GBIFMatch %>% 
  group_by(verbatimScientificName) %>% 
  mutate(verbatimNameDuplicate.flag = n() > 1) # TRUE if a duplicate
```

# Step 6: Preparation of output for non-plant taxa matched to GBIF

#### Add column with the source of the scientific name, the date accessed, and web-link

```{r}
GBIFMatch$scientificNameSource<-ifelse(GBIFMatch$unmatched.flag == TRUE, 
                                       "NA", paste0("GBIF", " ", "(", format(Sys.Date(), "%d %B %Y"),")", " ","https://www.gbif.org/species/", GBIFMatch$usageKey))
```

#### Remove unwanted columns

```{r}
GBIFMatch<-GBIFMatch %>% select(!(rank:usageKey)) # only keep the relevant columns
names(GBIFMatch) # names of the columns
```

# Step 7: Match plant names to BODATSA backbone

#### Merge plant dataset and BODATSA

Merge based on canonical name

```{r}
if(NameType == "canonical"){
BODATSAMatch<-
  left_join(PlantDat, BODATSA, by = join_by("verbatimScientificName" == "canonicalName"), relationship = "many-to-many")
}
```

Merge based on scientific name (with author information)

```{r}
if(NameType == "scientific"){
  BODATSA<- BODATSA %>% select(!(canonicalName)) # drop canonicalName from BODATSA
  BODATSAMatch<-
    left_join(PlantDat, BODATSA, by = join_by("verbatimScientificName" == "scientificName"), keep = TRUE, suffix = c('.1', '.2'))
}
```

#### Replace synonyms in "scientificName" column with accepted names from "Accepted" column

```{r}
BODATSAMatch<-BODATSAMatch %>%
  mutate(scientificName = replace(scientificName, is.na(Accepted)==FALSE, Accepted[is.na(Accepted)==FALSE]))
```

#### Identify taxa with multiple matches

```{r}
BODATSAMatch = BODATSAMatch %>% 
  group_by(verbatimScientificName) %>% 
  mutate(multipleMatches = n() > 1) # TRUE if there were multiple matches
```

#### Resolve, if possible, names for taxa with multiple matches

Take the match that is an accepted scientific name, if there is one

```{r}
matchesMult<-BODATSAMatch%>% filter(multipleMatches == TRUE & TaxStat == "acc") # create subset with multiple matches with accepted scientific names
matchesFilt<-BODATSAMatch%>% filter(!verbatimScientificName %in% matchesMult$verbatimScientificName) # create a subset without the above names
BODATSAMatch<-rbind(matchesMult, matchesFilt) # bind the subsets together
```

#### Address taxa with multiple matches that could not be resolved

Flag these taxa 

```{r}
BODATSAMatch = BODATSAMatch %>% 
  group_by(verbatimScientificName) %>% 
  mutate(multipleMatchesUnresolved = n() > 1) # TRUE if multiple matches are unresolved
```

Concatenate the multiple results for these taxa and separate with a pipe delimiter '|'

```{r}
BODATSAMatch<-BODATSAMatch %>%
  group_by(verbatimScientificName) %>%
  summarise(across(everything(), ~paste0(unique(.), collapse = "|")))
```

#### Write interim output to file 

```{r}
BODATSAInterim<-BODATSAMatch

write.csv(BODATSAInterim, "data//interim//BODATSAInterim.csv", row.names = FALSE)
```

# Step 8: Assess and flag issues for plants

#### Assess issues

Look at what results look like

```{r}
head(BODATSAMatch) # Look at the first 5 rows of the data
```

Look at how many have each status (e.g., accepted, synonym etc)

```{r}
BODATSAMatch %>%
  group_by(TaxStat) %>%
  count()
```

Look at how many were assigned accepted scientific names that had been corrected through fuzzy matching (Step 2a)  

```{r}
BODATSAMatch %>%
  group_by(synAccIssue) %>%
  count() # TRUE for taxa with this issue
```

Look at how many had multiple matches

```{r}
BODATSAMatch %>%
  group_by(multipleMatches) %>%
  count() # TRUE for taxa with multiple matches
```

Look at how many had multiple matches that were not resolved

```{r}
BODATSAMatch %>%
  group_by(multipleMatchesUnresolved) %>%
  count()  # TRUE for taxa with unresolved multiple matches
```

#### Flag taxonomic issues

Flag synonyms

```{r}
BODATSAMatch = BODATSAMatch %>% 
  mutate(synonym.flag = case_when((TaxStat =="syn") ~ TRUE, TRUE ~ FALSE)) # TRUE if a synonym
```

Flag taxon names that were not exact matches (i.e., accepted names corrected through fuzzy matching)

```{r}
BODATSAMatch = BODATSAMatch %>% 
  mutate(uncertain.flag=case_when((synAccIssue =="TRUE")~TRUE,
                                  TRUE~FALSE)) # TRUE if fuzzy matching involved
```

Flag doubtful taxa (uncertain, missapplied, and those which have multiple matches that are not resolved)

```{r}
BODATSAMatch <- BODATSAMatch %>%  mutate(doubtful.flag=case_when((multipleMatchesUnresolved =="TRUE")~TRUE,
                                                                  TRUE~FALSE)) %>% 
  mutate(doubtful.flag = str_detect(TaxStat, "unc|misap")) # TRUE if doubtful
```

Flag unmatched taxa

```{r}
BODATSAMatch = BODATSAMatch %>% 
  group_by(Genus) %>% 
  mutate(unmatched.flag = ifelse(Genus =="NA", TRUE, FALSE)) # TRUE if unmatched
```

Flag if the result should be checked for a possible error or issue (unmatched, doubtful, and uncertain matches)

```{r}
BODATSAMatch<-BODATSAMatch %>% 
  mutate(possibleError.flag = ifelse(uncertain.flag =="TRUE" | doubtful.flag == "TRUE" | unmatched.flag == "TRUE", TRUE, FALSE)) # TRUE if possible error
```

Flag sub-specific entities

```{r}
BODATSAMatch <- BODATSAMatch %>% 
  group_by(Rank1) %>% 
  mutate(subspecific.flag = ifelse(Rank1 =="NA", FALSE, TRUE)) # TRUE if sub-specific
```

Flag duplicates in returned scientific names (exclude NAs)

```{r}
BODATSAMatch = BODATSAMatch %>% 
  group_by(scientificName) %>% 
  mutate(scientificNameDuplicate.flag = ifelse(n() > 1 & scientificName !="NA", 
                                        TRUE , FALSE)) # TRUE if a duplicate
```

Flag duplicates in verbatim names

```{r}
BODATSAMatch = BODATSAMatch %>% 
  group_by(verbatimScientificName) %>% 
  mutate(verbatimNameDuplicate.flag = n() > 1) # TRUE if a duplicate
```

# Step 9: Extract unmatched plants

#### Create a subset of unmatched plants

```{r}
PlantUnMatched<-BODATSAMatch %>% filter(unmatched.flag == "TRUE") # subset of the unmatched plants
PlantUnMatched<-PlantUnMatched %>% select(verbatimScientificName:family) # remove unwanted columns
```

#### Create a subset of matched plants

```{r}
BODATSAMatch<-BODATSAMatch%>% filter(unmatched.flag == FALSE) # subset of matched plants
```

# Step 10: Preparation of output for plants matched to BODATSA 

#### Add column with the source of the scientific name, the date that BODATSA was downloaded, and a web-link

```{r}
BODATSAMatch$scientificNameSource<-paste0("BODATSA", " ", "(", BODATSADate,")"," ", "http://posa.sanbi.org/")
```

#### Remove unwanted columns

```{r}
BODATSAMatch <- BODATSAMatch %>% ungroup() # ungroup columns
BODATSAMatch<-BODATSAMatch %>% select(!(c(Family,Genus:multipleMatchesUnresolved))) # only keep the relevant columns
names(BODATSAMatch) # names of the columns
```

# Step 11: Match unmatched plant names to WCVP backbone

#### Match if verbatim names are canonical (without author information)

```{r}
if(NameType == "canonical"){
matches <- wcvp_match_names(PlantUnMatched,
                            name_col="verbatimScientificName",
                            author_col=NULL,
                            fuzzy=TRUE,
                            progress_bar=FALSE)
}
```

#### Match if verbatim names are scientific (with author information)

```{r}
if(NameType == "scientific"){
  matches <- wcvp_match_names(PlantUnMatched,
                              name_col="canonicalName",
                              author_col=NULL,
                              fuzzy=TRUE,
                              progress_bar=FALSE)
   }
```

#### Get the accepted scientific name for each taxon

```{r}
u<-which(wcvp_names$plant_name_id %in% matches$wcvp_accepted_id) # which rows in the WCVP dataset have the ids of the accepted taxon names
v <- wcvp_names[u,] %>% unite("scientificName", c(taxon_name, taxon_authors), na.rm = TRUE, sep = " ", remove = FALSE) # create a dataset of those names
matches<- left_join(matches, v, by = join_by("wcvp_accepted_id" == "plant_name_id")) # join that dataset to the matches dataset
matches<-matches%>% select((c(verbatimScientificName:wcvp_accepted_id, scientificName, powo_id))) # select useful columns
colnames(matches)[colnames(matches) == "family.x"] <- "family" # rename column
```

#### Resolve, if possible, taxon names with multiple matches

If the scientific name with author information is available take the match with that author

```{r}
if(NameType == "scientific"){
  matchesAuth<-matches%>% filter(multiple_matches == TRUE & match_type == "Exact (without author)") # multiple matches with exact matching
  matchesAuth<-matchesAuth %>%
    filter(str_detect(verbatimScientificName, fixed(wcvp_authors))) # detect authors in the verbatim name
  matchesAuth<-matchesAuth %>%
    mutate(match_type=replace(match_type, match_type=="Exact (without author)", "Exact")) # replace the match type with "Exact"
  matchesAuth<-matchesAuth %>%
    mutate(multiple_matches=replace(multiple_matches, multiple_matches==TRUE, FALSE)) # change multiple_matches to FALSE
    matches<-matches %>% 
    filter(!verbatimScientificName %in% matchesAuth$verbatimScientificName) # get a subset that doesn't contain resolved matches 
  matches<-rbind(matches, matchesAuth) # bind resolved matches to the subset
    }
```

For remaining taxa with multiple matches, select the match with an accepted scientific name, if there is one

```{r}
matchesMult<-matches%>% filter(multiple_matches == TRUE & wcvp_status == "Accepted") # subset of multiple matches, where there is an accepted taxon name
matchesFilt<-matches%>% filter(!verbatimScientificName %in% matchesMult$verbatimScientificName) # remove these taxa from the matches dataset
matches<-rbind(matchesMult, matchesFilt) # bind the two datasets together
```

#### Address taxa with multiple matches that could not be resolved

Flag these taxa 

```{r}
matches = matches %>% 
  group_by(verbatimScientificName) %>% 
  mutate(multiple_matchesUnresolved = n() > 1) # TRUE if multiple match could not be resolved
```

Concatenate the multiple results for these taxa and separate with pipe delimiter '|'

```{r}
WCVPMatch<-matches %>%
  group_by(verbatimScientificName) %>%
  summarise(across(everything(), ~paste0(unique(.), collapse = "|")))
```

#### Write interim output to file 

```{r}
WCVPInterim<- WCVPMatch

write.csv(WCVPInterim, "data//interim//WCVPInterim.csv", row.names = FALSE)
```

# Step 12: Assess and flag issues for plants matched to WCVP

#### Assess issues

```{r}
head(WCVPMatch) # Look at the first 5 rows of the data
```

Look at how many have each status (e.g., accepted, synonym etc)

```{r}
WCVPMatch %>%
  group_by(wcvp_status) %>%
  count()
```

Look at how many had a specific match type (e.g., exact, fuzzy etc)

```{r}
WCVPMatch %>%
  group_by(match_type) %>%
  count()
```

Look at how many fall into different ranks (e.g., species, subspecies etc)

```{r}
WCVPMatch %>%
  group_by(wcvp_rank) %>%
  count()
```

Look at how many had multiple matches

```{r}
WCVPMatch %>%
  group_by(multiple_matches) %>%
  count()
```

Look at how many had multiple matches that could not be resolved

```{r}
WCVPMatch %>%
  group_by(multiple_matchesUnresolved) %>%
  count()
```

#### Flag taxonomic issues

Flag synonyms

```{r}
WCVPMatch = WCVPMatch %>% 
  group_by(wcvp_status) %>% 
  mutate(synonym.flag = ifelse(wcvp_status =="Synonym", TRUE, FALSE)) # TRUE if a synonym
```

Flag taxon names that were not exact matches

```{r}
WCVPMatch<-WCVPMatch %>% 
  mutate(uncertain.flag = case_when((match_type =="NA")~FALSE,
                                    (match_type == "Exact")~FALSE,
                                    (match_type =="Exact (without author)") ~FALSE,
                                    TRUE~TRUE)) # TRUE if uncertain
```

Flag doubtful taxon names (e.g., invalid, illegitimate, missapplied, and those with unresolved multiple matches)

```{r}
WCVPMatch<-WCVPMatch %>% 
  mutate(doubtful.flag = case_when((multiple_matchesUnresolved == TRUE)~TRUE,
                                   (wcvp_status == "Accepted")~FALSE,
                                    (wcvp_status =="Synonym")~FALSE,
                                    (wcvp_status == "NA") ~FALSE,
                                    TRUE~TRUE)) # TRUE if doubtful
```

Flag unmatched taxa

```{r}
WCVPMatch = WCVPMatch %>% 
  mutate(unmatched.flag=ifelse(match_type =="NA", TRUE, FALSE)) # TRUE if not matched
```

Flag if the result should be checked for a possible error or issue (unmatched, doubtful, and uncertain matches)

```{r}
WCVPMatch<-WCVPMatch %>% 
  mutate(possibleError.flag = ifelse(uncertain.flag =="TRUE" | doubtful.flag == "TRUE" | unmatched.flag == "TRUE", TRUE, FALSE)) # TRUE if possible error or issue
```

Flag sub-specific entities

```{r}
WCVPMatch = WCVPMatch %>% 
  group_by(wcvp_rank) %>% 
  mutate(subspecific.flag = ifelse(wcvp_rank =="Species"|wcvp_rank == "NA", FALSE, TRUE)) # TRUE if sub-specific entity
```

Flag duplicates in returned scientific names (exclude NAs)

```{r}
WCVPMatch = WCVPMatch %>% 
  group_by(scientificName) %>% 
  mutate(scientificNameDuplicate.flag = ifelse(n() > 1 & scientificName !="NA", 
                                               TRUE , FALSE)) # TRUE if duplicate
```

Flag duplicates in verbatim names

```{r}
WCVPMatch = WCVPMatch %>% 
  group_by(verbatimScientificName) %>% 
  mutate(verbatimNameDuplicate.flag = n() > 1) # TRUE if duplicate
```

# Step 13: Preparation of output for plants matched to WCVP

#### Add column with the source of the scientific name, date accessed, and web-link

```{r}
WCVPMatch$scientificNameSource<-ifelse(WCVPMatch$unmatched.flag == "TRUE", 
                                       "NA", paste0("POWO", " ", "(", format(Sys.Date(), "%d %B %Y"),")", " ","https://powo.science.kew.org/taxon/urn:lsid:ipni.org:names:", WCVPMatch$wcvp_ipni_id))
```

#### Remove unwanted columns

```{r}
WCVPMatch <- WCVPMatch %>% ungroup() # ungroup columns
WCVPMatch <- WCVPMatch %>% select(!(c(match_type:wcvp_accepted_id, powo_id:multiple_matchesUnresolved))) # only keep the relevant columns
names(WCVPMatch) # names of the columns
```

# Step 14: Merge

#### Prepare the datasets

Standardise formating

```{r}
GBIFMatch<-as.data.frame(GBIFMatch)
BODATSAMatch<-as.data.frame(BODATSAMatch)
WCVPMatch<-as.data.frame(WCVPMatch)

GBIFMatch = GBIFMatch %>%
  mutate(across(everything(), as.character))
BODATSAMatch = BODATSAMatch %>%
  mutate(across(everything(), as.character))
WCVPMatch = WCVPMatch %>%
  mutate(across(everything(), as.character))
```

Revise flagged duplicate plant names (verbatim and scientific) so that they are across the BODATSA and WCVP matches

```{r}
PlantMatch<-rbind(BODATSAMatch, WCVPMatch) # Bind the datasets together
```

Flag duplicates in returned scientific names (exclude NAs)

```{r}
PlantMatch = PlantMatch %>% 
  group_by(scientificName) %>% 
  mutate(scientificNameDuplicate.flag = ifelse(n() > 1 & scientificName !="NA", 
                                                TRUE , FALSE)) # TRUE for duplicates
```

Flag duplicates in verbatim names

```{r}
PlantMatch = PlantMatch %>% 
  group_by(verbatimScientificName) %>% 
  mutate(verbatimNameDuplicate.flag = n() > 1) # TRUE for duplicates
```

#### Bind the datasets 

```{r}
AllMatch<-rbind(GBIFMatch, PlantMatch) # Bind all the datasets together
```

#### Check if the number of taxa in the output dataset is that same as in the original dataset

```{r}
length(AllMatch$verbatimScientificName) == length(NamesDat$verbatimScientificName) # should be TRUE
```

#### Reorder dataset, so that the taxa are in the same order as they were in the input file

```{r}
AllMatch<-AllMatch[match(NamesDat$verbatimScientificName, AllMatch$verbatimScientificName),]
```

#### Write final output to file

```{r}
write.csv(AllMatch, "data//processed//Taxon_names_standardised.csv", row.names = FALSE)
```
